Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array

### 2023-08-01 docs/manuscript/paper.tex
added Vargas plot
updates to compare

% The MNRAS class isn't designed to include a table of contents, but for this document one is useful.
% I therefore have to do some kludging to make it work without masses of blank space.
\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\section{Introduction}\label{sec:intro}
The inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Jaffe_2003, Wyithe2003,Sesana2013,McWilliams_2014,Ravi2015MNRAS.447.2772R,Burke2019, Skyes2022} is predicted to emit nHz gravitational waves (GWs). Other GW sources in this low-frequency regime include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}. The detection of nHz GWs has necessitated the development of new observational methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines. The foremost method is timing an ensemble of pulsars; a pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V}. A nHz GW influences the trajectory and frequency of radio pulses, leaving a characteristic impression on the pulse times of arrival (TOAs) measured at the  Earth. By measuring TOAs from multiple pulsars simultaneously one can effectively construct a detector with a baseline on the scale of parsecs. Multiple PTA detectors have been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{NANOgrav2023}, the Parkes Pulsar Timing array \citep[PPTA][]{Parkes2023}, and the European Pulsar Timing Array \citep[EPTA,][]{EPTA2023}. These individual efforts have joined in international collaboration, under the umbrella of the International Pulsar Timing Array \citep[IPTA,][]{2019MNRAS.490.4666P}, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}, MeerTime \citep{meertime2,Meertime} and the Chinese PTA \citep{Hobbs_2019}. \newline

\textcolor{red}{TK: Content and references need to be updated in light of recent PTA results}. The incoherent superposition of multiple SMBHB sources leads to a stochastic GW background detectable at nHz frequencies \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}. Previous efforts have mainly focused on detecting the stochastic background by measuring the cross-correlation between the pulsar timing residuals between pairs of pulsars as a function of the angular separation between the pulsars -- the Hellings-Downs curve \citep{Hellings}. However, no GW background has yet been detected \citep{Lentati2015,NanoGrav2018,2022MNRAS.510.4873A}. \newline

Individual SMBHBs that are sufficiently massive and nearby may be resolvable with PTAs, allowing the very earliest stages of their evolution and coalescence to be investigated \citep{Sesana2010,Yardley2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}. 
Indeed, the stochastic GW background itself may be dominated by a few individual binary sources \citep{Ravi2012singlesource}. Individual SMBHBs are continuous wave sources: they generate persistent, quasi-monochromatic modulations of a known form in pulsar timing residuals. Consequently, they are detected more efficiently by either a frequentist matched filter e.g.\ the ${\cal F}$-statistic \citep{Lee2011MNRAS.414.3251L, Ellis2012ApJ,Zhu2014PPTA} or else Bayesian inference \citep{Ellis2016,Arzoumanian2020A} rather than by cross-correlating pulsar pairs. However, PTA observational efforts to detect individual sources have thus far been unsuccessful \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}. \newline

@@

The full set of Kalman recursions is presented in Appendix \ref{sec:kalman}. At each discrete timestep $i = 1, ... , M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i$, and the covariance of those estimates $\boldsymbol{P}_i$. The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by projecting the state predictions back into measurement space, $\hat{\boldsymbol{X}}_i \mapsto \hat{\boldsymbol{Y}}_i$, via the measurement equation, Eq. \ref{eq:measurement}d. The measurement predictions can then be compared aganst the true observed noisy measurements to get a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, sometimes termed the ``innovation". The Kalman filter also calculates the uncertainty in $\boldsymbol{\epsilon}_i$ via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{\intercal} \rangle$. The Gaussian log-likelihood can then be calculated at each timestep,
\begin{eqnarray}
	\log \mathcal{L}_i =  -\frac{1}{2} \left (N \log 2 \pi + \log  \left | \boldsymbol{S}_i \right | + \boldsymbol{\epsilon}_i^{\intercal} \boldsymbol{S}_i^{-1}  \boldsymbol{\epsilon}_i \right )
\end{eqnarray}
with the total log-likelihood the sum over all timesteps, i.e. 
\begin{eqnarray}
	\log \mathcal{L} =  \sum_{i=1}^{i=M} \log \mathcal{L}_i  \label{eq:likelihood}
\end{eqnarray}
The likelihood $\mathcal{L}$, along $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are functions of the static parameters of the model, $\bar{\theta}$. If the estimates of the system parameters, $\hat{\theta}$, that we pass to the Kalman filter are close to the true underlying parameters then the error in $\hat{\boldsymbol{X}}$, $\hat{\boldsymbol{Y}}$ is minimized and $\mathcal{L}$ is maximised. This is illustrated in Figure \ref{fig:6MB_BFS}; given a data timeseries of the measured pulsar frequency the Kalman filter is able to recover the hidden state evolution with high fidelity - leavign only gaussian errrs

@@

\section{Tests with synthetic data} \label{sec:testing}
We go on to discuss our choice of pulsars to make up our PTA in Section \ref{sec:pta_pulsars}, before deploying these techniques in Sections \ref{sec:param_estimation} , \ref{sec:detection} for parameter estimation and model selection.

\subsection{Synthetic data generation}

@@

For this work we generate representative synthetic data as follows. We first specify a set of pulsars to make up our artificial PTA. We take the 47 pulsars that make up the 12.5 year NANOGrav PTA \citep{2020ApJ...905L..34A}. For each pulsar we get the sky location (which defines the vector $\boldsymbol{q}^{n}$), the distance, the barycentric rotation frequency - which we identify with $f_{\rm em}^{(n)} (t_1)$ -  and the time derivative of the the barycentric rotation frequency - which we identify with $\dot{f}_{\rm em}^{(n)} (t_1)$. The spatial distribution and the distances of the pulsars used for this artificial PTA are shown in Fig \ref{fig:pulsar_distrib}. The data is obtained via the Australia Telescope National Facility pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}. The remaining unspecified pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$. We disused in Section \ref{sec:psr_frequency} how the quantity $[\sigma^{(n)}]^2 /\gamma^{(n)} $ sets the amplitude of the characteristic root mean square fluctuations in $f_p^{(n)}(t)$. Typically the mean reversion timescale $[\gamma^{(n)}]^{-1} >> T_{\rm obs}$ \cite{Vargas}, whilst $[\sigma^{(n)}]^2$ varies between individual pulsars. For this work we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$. In order to set a physically reasonable value for $[\sigma^{(n)}]^2$ we use two complementary approaches. The first approach uses the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} where the standard deviation in the pulsar TOA is given by
\begin{eqnarray}
	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)= \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right)+ \alpha_4 \ln \left( \frac{T}{ 1 \text{ year}} \right) \label{eq:sigmap}
\end{eqnarray}
where $T$ is the length of the observation span and for MSPs $\ln \alpha_1 \sim -20 $, $\alpha_2 \sim 1$, $\alpha_3 \sim 2$ and $\alpha_4 \sim 2.4$. Throughout this work we assume that all pulsars are observed with a weekly cadence i.e. $T = 1$ week. The frequency noise parameter can then be calculated for each pulsar where the standard deviation in the pulsar TOA can be related to the standard deviation in the pulsar frequency as 
\begin{eqnarray}
	\sigma^{(n)} = \sigma_{\rm TOA}^{(n)} \frac{f_{\rm p}^{(n)}}{T} \ . \label{eq:sigmap_f}
\end{eqnarray}
For our synthetic NANOGrav PTA, the median $\sigma^{(n)} = 4.28 \times 10^{-21} $ s. Alternatively to this empirical best-fit model, we can also directly solve the state equation Eq. \ref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}} so as to obtain a phase solution
\begin{eqnarray}
	\phi^{(n)}(t) = \int_0^t dt' f^{(n)}_{\rm p}(t') 
\end{eqnarray}
We can then vary $\sigma^{(n)}$ and compare the synthetically generate phase residuals with the phase solution from the UTMOST pulsar timing program with the Mongolo Observatory Synthesis Telescope  \citep{UTMOST}. It is sufficient to simply compare the synthetic and real solutions visually, rather than e.g. defining some quantitative loss function (e.g. mean-squared error) between the two solutions and solving the optimisation problem of the value of $\sigma_{\rm TOA}^{(n)}$ which minimised this loss function; we emphasise that we are not overly concerned with getting the maximally accurate values for the noise in the pulsar frequency evolution, but just instead some reasonable, representative values. This approach of generating synthetic phase solutions and visually comparing with real solutions is also followed in \cite{Vargas}. In Figure \ref{fig:qualitative_compare} we demonstrate for a particular representative pulsar, PSR XXX, blah blah blah...
\begin{figure}
	\includegraphics[width=\columnwidth]{example-image-a}
	\caption{Actual and synthetic phase evolution}
	\label{fig:qualitative_compare}
\end{figure}

\subsubsection{Synthetic data generation}
We generate $N$ synthetic noisy timeseries by integrating the state equations, Eq \ref{eq:spinevol}- \ref{eq:xieqn} numerically, mapping from state space to measurement space via the measurement equation, Eq. \ref{eq:measurement}, and finally adding to each timeseries some zero-mean Gaussian noise with covariance $\sigma_{\rm m}^2$. We take $\sigma_{\rm m}^2$ to be known \textit{a priori} rather than a parameter to be inferred. Throughout this work we fix $\sigma_{\rm m}^2 = 10^{-11}$. Equations \ref{eq:spinevol}- \ref{eq:xieqn} are solved by a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}. The static pulsar parameters  $\boldsymbol{\theta}_{\rm psr}$ are completely specified by our construction of the synthetic PTA outlined in Section \ref{sec:synt_pta}. For this work we consider all pulsars to be observed for $T_{\rm obs} =10$ years, uniformly sampled with a weekly cadence. In practice, in order to avoid numerical precision issues that arise when solving Eq \ref{eq:spinevol}- \ref{eq:xieqn} $\left ( \text{since } \sigma^{(n)} \ll f_{\rm p}^{(n)} \right)$ we first ``heterodyne" our state by subtracting the deterministic evolution, equivalent to a change of variables:
\begin{equation}
	f_{\rm p}^{\prime (n)} = f_{\rm p}^{(n)} - f_{\rm em}^{(n)}
\end{equation}  
We similarly heterodyne the measurement variable 
\begin{equation}
	f_{\rm m}^{\prime (n)} = f_{\rm m}^{(n)} - f_{\rm em}^{*(n)}
\end{equation}
where $ f_{\rm em}^{*(n)}$ is a guess of the deterministic spin-down based on the pulsar ephemeris. For synthetic data we can set $ f_{\rm em}^{*(n)} = f_{\rm em}^{(n)}$ but this is not true generally for real-world observations since the spin-down ephemeris is only known approximately, not exactly. The measurement equation Eq. \ref{eq:measurement} that relates the state and measurement variables is then updated as 
\begin{equation}
	f_{\rm m}^{\prime (n)}(t) = f_{\rm p}^{\prime (n)}(t-d) g^{(n)}(t) -  f_{\rm em}^{(n)}((t-d))(1-g^{(n)}(t))
	\label{eq:measurement_cov}
\end{equation}
We emphasise that this change of variables is simply a convenient scaling to bring our numerical values into a reasonable dynamic range without having to use extended numerical floating point formats (e.g. long double, quadruple). It does not remove any degrees of freedom of the formulation. In particular both $f_{\rm em}^{(n)}(t_1)$
and $\dot{f}_{\rm em}^{(n)}(t_1)$ remain static parameters of the model, but now appear in the measurement equation rather than the state equation.

@@

\subsubsection{Parameter estimation}\label{sec:parameter_estim}
From Eq. \ref{eq:g_func_trig} it can be seen that the measurement equation generally separates into two cosine terms. The first term $\cos(-\Omega t + \Phi_0)$ depends only on the GW source parameters and is shared across all pulsars. The argument of the cosine corresponds to the GW phase at the observer on Earth.  Conversely, the second term, $\cos \left(-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d \right)$ varies between pulsars. The argument of the cosine corresponds to the GW phase at each the individual pulsar. The first and second terms are commonly referred to as the ``Earth term" and ``pulsar term" respectively. Whilst the Earth term is phase coherent between all pulsars and so can be summed across the array to increase the total signal-to-noise, the pulsar terms each have individual phases and in standard PTA analysis are typically considered as a source of self noise and dropped from the analysis \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C} at the expense of a modestly reduced detection probability ($\sim 5 \%$) and the introduction of a bias in the sky localisation \citep{Zhupulsarterms,Chen2022}. In this work we follow the standard analysis approach and drop the pulsar terms from our model. We defer the inclusion of the pulsar terms in the measurement equation to a future work. We stress that the pulsar terms are dropped only for purposes of Bayesian inference, i.e. from the Kalman filter model that feeds into the nested sampling algorithm. The synthetic data that we apply our method on does include the pulsar term in full. \newline

First we consider a single noise realisation of the synthetic data for the representative example. We apply the Kalman filter using the earth terms only measurement equation in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters, which are shown in Figure X.

@@

\subsection{Practical considerations for inference}

heterodyeing

pulsar terms vs earth terms

For this initial demonstrative work we use only the Earth terms

loss of SNR, cite Zhu etc.

We defer the inclusion of the pulsar terms for a future work

reduces the dimensionality of the problem to $7 + 4N$ since the measurement equation eq X is no longer a function of the pulsar distance.

we emphasise that whilst our alternative model $m$ disregards the pulsar terms, the synthetic data does have the pulsar terms

\subsection{PTA pulsars}\label{sec:pta_pulsars}

\noindent With our Kalman filter and nested sampling techniques in hand,

\noindent Within our state-space formulation, the pulsar evolution is governed by a set of 5 parameters, $\bar{\theta}_{\rm PSR}$ for each pulsar. The parameters $f_{\rm EM} (0) $,$ \dot{f}_{\rm EM} (0)$ and $d$ are well specified via existing pulsar datasets. We take the frequency and frequency derivative as returned from the pulsar datasets to simply be the values now at $t=0$. The pulsar distances are also known, though less well constrained. Going forward we take the distances retuned from the datasets as the true values of the pulsars that make up our synthetic PTA. The specification of $\gamma$ and $\sigma$ are more involved. $\gamma$ specifies an effective timescale of reversion to the mean \textcolor{red}{TK: Need some discussion and though on how to choose these two parameters to be astrophysically reasonable c.f. A. Vargas}

we consider the pulsars to have data over the entire

\subsection{Parameter estimation}\label{sec:param_estimation}

>>>>

% The MNRAS class isn't designed to include a table of contents, but for this document one is useful.
% I therefore have to do some kludging to make it work without masses of blank space.
\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\section{Introduction}\label{sec:intro}
The inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Jaffe_2003, Wyithe2003,Sesana2013,McWilliams_2014,Ravi2015MNRAS.447.2772R,Burke2019, Skyes2022} is predicted to emit nHz gravitational waves (GWs). Other GW sources in this low-frequency regime include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}. The detection of nHz GWs has necessitated the development of new observational methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines. The foremost method is timing an ensemble of pulsars; a pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V}. A nHz GW influences the trajectory and frequency of radio pulses, leaving a characteristic impression on the pulse times of arrival (TOAs) measured at the  Earth. By measuring TOAs from multiple pulsars simultaneously one can effectively construct a detector with a baseline on the scale of parsecs. Multiple PTA detectors have been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{NANOgrav2023}, the Parkes Pulsar Timing array \citep[PPTA,][]{Parkes2023}, and the European Pulsar Timing Array \citep[EPTA,][]{EPTA2023}. These individual efforts have joined in international collaboration, under the umbrella of the International Pulsar Timing Array \citep[IPTA,][]{2019MNRAS.490.4666P}, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}, MeerTime \citep{meertime2,Meertime} and the Chinese PTA \citep[CPTA,][]{Hobbs_2019}. \newline

The incoherent superposition of multiple SMBHB sources leads to a stochastic GW background detectable at nHz frequencies \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}. Previous efforts have mainly focused on detecting the stochastic background by measuring the cross-correlation between the pulsar timing residuals between pairs of pulsars as a function of the angular separation between the pulsars -- the Hellings-Downs curve \citep{Hellings}. After multiple non-detections \citep{Lentati2015,NanoGrav2018,2022MNRAS.510.4873A} consilient evidence for the GW background was presented by NANOGrav \citep{2023ApJ...951L...8A}, EPTA/InPTA \citep{2023arXiv230616214A}, PPTA \citep{2023ApJ...951L...6R} and the CPTA \citep{2023RAA....23g5024X} \newline

Individual SMBHBs that are sufficiently massive and nearby may be resolvable with PTAs, allowing the very earliest stages of their evolution and coalescence to be investigated \citep{Sesana2010,Yardley2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}. 
Indeed, the stochastic GW background itself may be dominated by a few individual binary sources \citep{Ravi2012singlesource}. Individual SMBHBs are continuous wave sources: they generate persistent, quasi-monochromatic modulations of a known form in pulsar timing residuals. Consequently, they are detected more efficiently by either a frequentist matched filter e.g.\ the ${\cal F}$-statistic \citep{Lee2011MNRAS.414.3251L, Ellis2012ApJ,Zhu2014PPTA} or else Bayesian inference \citep{Ellis2016,Arzoumanian2020A} rather than by cross-correlating pulsar pairs. However, PTA observational efforts to detect individual sources have thus far been unsuccessful \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}, with inconclusive evidence at low significance presented by the EPTA for a individual source at 4-5 nHz \citep{2023arXiv230616226A}. \newline

@@

The full set of Kalman recursions is presented in Appendix \ref{sec:kalman}. At each discrete timestep $i = 1, ... , M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i$, and the covariance of those estimates $\boldsymbol{P}_i$. The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by projecting the state predictions back into measurement space, $\hat{\boldsymbol{X}}_i \mapsto \hat{\boldsymbol{Y}}_i$, via the measurement equation, Eq. \ref{eq:measurement}. The measurement predictions can then be compared aganst the true observed noisy measurements to get a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, sometimes termed the ``innovation". The Kalman filter also calculates the uncertainty in $\boldsymbol{\epsilon}_i$ via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{\intercal} \rangle$. The Gaussian log-likelihood can then be calculated at each timestep,
\begin{eqnarray}
	\log \mathcal{L}_i =  -\frac{1}{2} \left (N \log 2 \pi + \log  \left | \boldsymbol{S}_i \right | + \boldsymbol{\epsilon}_i^{\intercal} \boldsymbol{S}_i^{-1}  \boldsymbol{\epsilon}_i \right )
\end{eqnarray}
with the total log-likelihood the sum over all timesteps, i.e. 
\begin{eqnarray}
	\log \mathcal{L} =  \sum_{i=1}^{i=M} \log \mathcal{L}_i  \label{eq:likelihood}
\end{eqnarray}
The likelihood $\mathcal{L}$, along $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are functions of the static parameters of the model, $\bar{\theta}$. If the estimates of the system parameters, $\hat{\theta}$, that we pass to the Kalman filter are close to the true underlying parameters then the error in $\hat{\boldsymbol{X}}$, $\hat{\boldsymbol{Y}}$ is minimized and $\mathcal{L}$ is maximised. This is illustrated in Figure \ref{fig:6MB_BFS}; given a data timeseries of the measured pulsar frequency the Kalman filter is able to recover the hidden state evolution with high fidelity - leavign only gaussian errrs

@@

\section{Tests with synthetic data} \label{sec:testing}

In this section we use synthetic data to test the ability of our method to detect the presence of a GW in noisy data and recover the static parameters $\boldsymbol{\theta}$. In Section \ref{sec:synth_data} we discuss how the synthetic data is generated, including the construction of an artificial PTA. In Section \ref{sec:rep_example} we generate synthetic data for a single representative example SMBHB system and apply our method to recover the system parameters and the Bayes factor. \textcolor{red}{TK: more here once this section is completed}.

\subsection{Synthetic data generation} \label{sec:synth_data}

@@

For this work we generate representative synthetic data as follows. We first specify a set of pulsars to make up our artificial PTA. We take the 47 pulsars that make up the 12.5 year NANOGrav PTA \citep{2020ApJ...905L..34A}.

For each pulsar we get the sky location (which defines the vector $\boldsymbol{q}^{n}$), the distance, the barycentric rotation frequency - which we identify with $f_{\rm em}^{(n)} (t_1)$ -  and the time derivative of the the barycentric rotation frequency - which we identify with $\dot{f}_{\rm em}^{(n)} (t_1)$. The spatial distribution and the distances of the pulsars used for this artificial PTA are shown in Fig \ref{fig:pulsar_distrib}. The data is obtained via the Australia Telescope National Facility pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}. The remaining unspecified pulsar parameters are $\gamma^{(n)}$ and $\sigma^{(n)}$. We disused in Section \ref{sec:psr_frequency} how the quantity $[\sigma^{(n)}]^2 /\gamma^{(n)} $ sets the amplitude of the characteristic root mean square fluctuations in $f_p^{(n)}(t)$. Typically the mean reversion timescale $[\gamma^{(n)}]^{-1} >> T_{\rm obs}$ \cite{Vargas}, whilst $[\sigma^{(n)}]^2$ varies between individual pulsars. For this work we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$. In order to set a physically reasonable value for $[\sigma^{(n)}]^2$ we use two complementary approaches. The first approach uses the empirical timing noise model from \cite{Shannon2010ApJ...725.1607S} where the standard deviation in the pulsar TOA is given by
\begin{eqnarray}
	\ln \left(\frac{\sigma_{\rm TOA}^{(n)}}{\mu s} \right)= \ln \alpha_1 +  \alpha_2 \ln f_{\rm p}^{(n)} + \alpha_3 \ln \left(\frac{\dot{f}_{\rm p}^{(n)}}{10^{-15} \text{s}^{-2}} \right)+ \alpha_4 \ln \left( \frac{T}{ 1 \text{ year}} \right) \label{eq:sigmap}
\end{eqnarray}
where $T$ is the length of the observation span and for MSPs $\ln \alpha_1 \sim -20 $, $\alpha_2 \sim 1$, $\alpha_3 \sim 2$ and $\alpha_4 \sim 2.4$. Throughout this work we assume that all pulsars are observed with a weekly cadence i.e. $T = 1$ week. The frequency noise parameter can then be calculated for each pulsar where the standard deviation in the pulsar TOA can be related to the standard deviation in the pulsar frequency as 
\begin{eqnarray}
	\sigma^{(n)} = \sigma_{\rm TOA}^{(n)} \frac{f_{\rm p}^{(n)}}{T} \ . \label{eq:sigmap_f}
\end{eqnarray}
For our synthetic NANOGrav PTA, the median $\sigma^{(n)} = 4.28 \times 10^{-21} $ s. Alternatively to this empirical best-fit model, we can also directly solve the state equation Eq. \ref{eq:spinevol} numerically using the \texttt{baboo} package \footnote{\url{https://github.com/meyers-academic/baboo}} so as to obtain a phase solution
\begin{eqnarray}
	\phi^{(n)}(t) = \int_0^t dt' f^{(n)}_{\rm p}(t') 
\end{eqnarray}
For a particular choice of $\gamma^{(n)}$ we can then vary $\sigma^{(n)}$ and compare the synthetically generated phase residuals with the empirical phase residuals. We obtain empirical phase residuals from the NANOGrav 12.5 year narrowband data set \cite{nanograv_narrowband_DR}. For the purposes of validating the $\sigma^{(n)}$ inferred from Eq \ref{eq:sigmap_f}, it is sufficient at this stage to simply compare the synthetic and real solutions visually. Whilst it would be possible to e.g. define some quantitative loss function (e.g. mean-squared error) between the two solutions and solve the optimisation problem to infer $\sigma^{(n)}$, for our purposes this is overkill and it is satisfactory to simply qualitatively compare the two solutions as a useful sanity check. We emphasise that we are not overly concerned with getting the maximally accurate values for $\sigma^{(n)}$, but just instead some reasonable values for constructing representative synthetic data. This approach of generating synthetic phase solutions and visually comparing with empirical solutions is also followed in \cite{Vargas}. In Figure \ref{fig:qualitative_compare} we compare the synthetic and empirical residuals for 3 NANOgrav pulsars: J0030+0451, J0340+4130 and J0613-0200. We can see that for all pulsars the synthetic and empirical residuals are qualitatively similar with the magnitude of the residuals being directly comparable for each pulsar. 
   \begin{figure}
	\subfloat[PSR J0030+0451]{{\includegraphics[width=\columnwidth]{images/example_residuals_plot} }\label{fig:mycaption-a}}

	\subfloat[PSR J0340+4130 ]{{\includegraphics[width=\columnwidth]{images/example_residuals_plot} }\label{fig:mycaption-b}}

	\subfloat[PSR J0613-0200]{{\includegraphics[width=\columnwidth]{images/example_residuals_plot} }\label{fig:mycaption-c}}
	\caption{Actual (left panels) and synthetic (right panels) phase residuals for 3 NANOGrav pulsars. The actual residuals are obtained via the NANOGrav 12.5 year data release \citep{nanograv_narrowband_DR}. The synthetic residuals are generated by numerically solving Eq. \ref{eq:spinevol} with $\gamma^{(n)} = 10^{-13}$ and $\sigma^{(n)}$ inferred from Eqs. \ref{eq:sigmap}, \ref{eq:sigmap_f}. For all pulsars the two solutions are qualitatively similar, with comparable magnitudes in the residuals. \textcolor{red}{TK: these plots need updating, but need to discuss with Andres re use of baboo}}
	\label{fig:qualitative_compare}
\end{figure}

%
%\textbf{phase solution from the UTMOST pulsar timing program with the Mongolo Observatory Synthesis Telescope  \citep{UTMOST}. I}
%

\subsubsection{Synthetic data generation}
We generate $N$ synthetic noisy timeseries by integrating the state equations, Eq \ref{eq:spinevol}- \ref{eq:xieqn} numerically for the synthetic PTA described in Sectin \ref{sec:synt_pta}, mapping from state space to measurement space via the measurement equation, Eq. \ref{eq:measurement}, and finally adding to each timeseries some zero-mean Gaussian noise with covariance $\sigma_{\rm m}^2$. Equations \ref{eq:spinevol}- \ref{eq:xieqn} are solved by a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}. The static pulsar parameters  $\boldsymbol{\theta}_{\rm psr}$ are completely specified by our construction of the synthetic PTA outlined in Section \ref{sec:synt_pta}. For this work we consider all pulsars to be observed for $T_{\rm obs} =10$ years, uniformly sampled with a weekly cadence. We take $\sigma_{\rm m}^2$ to be known \textit{a priori} rather than a parameter to be inferred. The measurement noise can be approximately related to the uncertainty in the pulse TOA $\sigma_{\rm TOA}$ as
\begin{equation}
	\sigma_{\rm m} = f_{\rm p}^{(n)} \frac{\sigma_{\rm TOA}}{\text{cadence}} \ . 
\end{equation}
For a millisecond pulsar with $f_{\rm p}^{(n)}$ observed with a weekly cadence and $\sigma_{\rm TOA} \sim 1 \mu$ s this gives $\sigma_{\rm m} \sim 10^{-10}$, whilst the very best pulsars might have $\sigma_{\rm TOA} \sim 10 $ ns or $\sigma_{\rm m} \sim 10^{-12}$. Throughout this work we fix $\sigma_{\rm m} = 10^{-11}$. In practice, in order to avoid numerical precision issues that arise when solving Eq \ref{eq:spinevol}- \ref{eq:xieqn} $\left ( \text{since } \sigma^{(n)} \ll f_{\rm p}^{(n)} \right)$ we first ``heterodyne" our state by subtracting the deterministic frequency evolution, equivalent to a change of variables:
\begin{equation}
	f_{\rm p}^{\prime (n)} = f_{\rm p}^{(n)} - f_{\rm em}^{(n)}
\end{equation}  
We similarly heterodyne the measurement variable 
\begin{equation}
	f_{\rm m}^{\prime (n)} = f_{\rm m}^{(n)} - f_{\rm em}^{*(n)}
\end{equation}
where $ f_{\rm em}^{*(n)}$ is a guess of the deterministic spin-down based on the pulsar ephemeris. For synthetic data we can set $ f_{\rm em}^{*(n)} = f_{\rm em}^{(n)}$ but this is not true generally for real-world observations since the spin-down ephemeris is only known approximately, not exactly. The measurement equation Eq. \ref{eq:measurement} that relates the state and measurement variables is then updated as 
\begin{equation}
	f_{\rm m}^{\prime (n)}(t) = f_{\rm p}^{\prime (n)}(t-d) g^{(n)}(t) -  f_{\rm em}^{(n)}((t-d))(1-g^{(n)}(t))
	\label{eq:measurement_cov}
\end{equation}
We emphasise that this change of variables is simply a convenient scaling to bring our numerical values into a reasonable dynamic range without having to use extended numerical floating point formats (e.g. long double, quadruple). It does not remove any degrees of freedom of the formulation. In particular both $f_{\rm em}^{(n)}(t_1)$
and $\dot{f}_{\rm em}^{(n)}(t_1)$ remain static parameters of the model, but now appear in the measurement equation rather than the state equation.

@@

\subsubsection{Parameter estimation}\label{sec:parameter_estim}
From Eq. \ref{eq:g_func_trig} it can be seen that the measurement equation generally separates into two cosine terms. The first term $\cos(-\Omega t + \Phi_0)$ depends only on the GW source parameters and is shared across all pulsars. The argument of the cosine corresponds to the GW phase at the observer on Earth.  Conversely, the second term, $\cos \left(-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d \right)$ varies between pulsars. The argument of the cosine corresponds to the GW phase at each the individual pulsar. The first and second terms are commonly referred to as the ``Earth term" and ``pulsar term" respectively. Whilst the Earth term is phase coherent between all pulsars and so can be summed across the array to increase the total signal-to-noise, the pulsar terms each have individual phases and in standard PTA analysis are typically considered as a source of self noise and dropped from the analysis \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C} at the expense of a modestly reduced detection probability ($\sim 5 \%$) and the introduction of a bias in the sky localisation \citep{Zhupulsarterms,Chen2022}. In this work we follow the standard analysis approach and drop the pulsar terms from our model. We defer the inclusion of the pulsar terms in the measurement equation to a future work. This choice reduces the dimensionality of the parameter estimation problem to $7 + 4N$ since the measurement equation Eq \ref{eq:measurement} is no longer a function of the pulsar distance.  We stress that the pulsar terms are dropped only for purposes of Bayesian inference, i.e. from the Kalman filter model that feeds into the nested sampling algorithm. The generated synthetic data that we use to test our method does include the pulsar term in full. \newline

Initially we consider a single noise realisation of the synthetic data for the representative example. We apply the Kalman filter using the earth terms only measurement equation in conjunction with nested sampling in order to infer the posterior distributions in each of the parameters, which are shown in Figure \ref{fig:corner_plot_1}.

the gw parameters....

the frequency parameters....

the sigma p parameters....

bias in parameters c.f. likelihood

@@

\subsection{Multiple noise realisations}

The above discussion was just for a single realisaitn of the noise

\subsubsection{Exploring a broader parameter space}

The above was for a particular SBHB system

hellings downs

multiple sources
